---
title: "Forecasting Oman Oil Prices Using Statical Models"
author: "Abdullah"
format: pdf
editor: visual
---

# General Introduction

Oil is one of the most important natural resources in the world and plays a key role in the economic development of many countries. In the Sultanate of Oman, oil is a major source of national income and contributes significantly to government revenue. Because of this, changes in oil prices can strongly affect the country’s economy, including budgeting, investment, and development planning.

Oil prices are known to be highly volatile over time. They are influenced by many factors such as global demand and supply, political events, economic conditions, and decisions made by oil-producing countries. For example, events like economic crises or changes in production levels can cause oil prices to rise or fall sharply. This volatility makes it difficult for governments and businesses to predict future prices accurately.

To address this challenge, different forecasting methods have been developed. Traditional statistical methods, especially time series analysis, are widely used to study data that changes over time. Models such as Autoregressive Integrated Moving Average (ARIMA) and Exponential Smoothing help to identify patterns like trends and seasonality in oil prices and use these patterns to make forecasts.

In recent years, machine learning techniques have become popular for prediction problems. These methods, including regression models, Random Forest, and Support Vector Regression, can handle complex relationships in data and may improve forecasting accuracy. Unlike traditional models, machine learning approaches can capture non-linear patterns more effectively.

This study aims to analyze oil price data in Oman using both time series and machine learning techniques. The purpose is to compare the performance of these methods and determine which approach provides more accurate forecasts. By doing so, the study will provide useful insights into oil price behavior and support better decision-making.

In conclusion, accurate forecasting of oil prices is important for economic planning in Oman, and this study contributes by applying and comparing different analytical approaches.

# Importance of Oil in Oman Economy

Oil plays a central role in the economy of the Sultanate of Oman and is considered the backbone of the country’s economic development. Since the discovery of oil in Oman in the 1960s, the oil sector has been the main source of national income and government revenue. It contributes significantly to the country’s Gross Domestic Product (GDP), exports, and public finances.

One of the most important roles of oil in Oman is its contribution to government revenue. A large portion of the government’s budget is financed through oil exports. This revenue is used to support public services such as education, healthcare, infrastructure, and social development programs. Therefore, any change in oil prices can directly affect government spending and economic stability.

In addition, oil exports represent a major share of Oman’s total exports. The country relies heavily on exporting crude oil to international markets, making it an important participant in the global energy market. High oil prices generally lead to increased export earnings, while low prices can reduce national income and create economic challenges.

The oil sector also supports economic growth by creating employment opportunities and encouraging investment. It has helped in the development of related industries such as transportation, construction, and energy services. Furthermore, oil revenues have been used to invest in diversification projects aimed at reducing dependence on oil in the long term.

However, the strong reliance on oil also makes Oman’s economy vulnerable to fluctuations in global oil prices. Sudden drops in oil prices can lead to budget deficits and economic uncertainty. For this reason, accurate analysis and forecasting of oil prices are very important for effective economic planning and policy-making.

In conclusion, oil remains a vital component of Oman’s economy, influencing its growth, stability, and development. Understanding oil price movements is therefore essential for ensuring sustainable economic progress.

# Research questions

1.  What are the main patterns in Oman oil prices over time?

2.  Can time series models accurately forecast Oman oil prices?

3.  Can machine learning techniques improve prediction accuracy?

4.  Which method performs better for forecasting Oman oil prices?

# Database

```{r}
# Install packages if needed
library(readxl)
library(ggplot2)
library(psych)
library(corrplot)
library(lubridate)

library(dplyr)
library(tidyr)
library(zoo)
library(lubridate)
library(writexl)
```

# 1. Read the Excel file

```{r}
df <- read_excel("Data_abdullah.xlsx")
head(df)

```

#Create a proper monthly date column

```{r}
df <- df %>%
  mutate(
    years = as.integer(years),
    month = factor(month, levels = month.name, ordered = TRUE),
    date = as.Date(paste(years, month, 1), format = "%Y %B %d")
  ) %>%
  arrange(date)

```

#2.Data Cleaning and Preparation #Check missing values

```{r}
cat("Missing values before imputation:\n")
print(colSums(is.na(df)))

cat("\nRows with missing values:\n")
print(df %>% filter(if_any(everything(), is.na)))


ggplot(df) +
  geom_line(aes(x = 1:nrow(df), y = prices, color = "Oil Prices")) +
  geom_line(aes(x = 1:nrow(df), y = exports, color = "Oil Exports")) +
  geom_line(aes(x = 1:nrow(df), y = productions, color = "Production")) +
  geom_line(aes(x = 1:nrow(df), y = USD_to_OMR, color = "Exchange Rate")) +
  labs(title = "Trend of Oil Variables",
       x = "Time",
       y = "Value") +
  scale_color_manual(values=c("blue","red","orange","black"))
```

#Impute missing values based on the monthly average from all years

```{r}

# First, create a 'month' column to group the data by month (if not already present)
df$month <- factor(df$month, levels = month.name, ordered = TRUE)

# Impute missing values based on the monthly average from all years
df_imputed <- df %>%
  group_by(month) %>%
  mutate(across(c(prices, exports, productions, USD_to_OMR), 
                ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  ungroup()

# Check the result
print(head(df_imputed))

cat("\nMissing values after imputation:\n")
print(colSums(is.na(df_imputed)))

ggplot(df_imputed) +
  geom_line(aes(x = 1:nrow(df), y = prices, color = "Oil Prices")) +
  geom_line(aes(x = 1:nrow(df), y = exports, color = "Oil Exports")) +
  geom_line(aes(x = 1:nrow(df), y = productions, color = "Production")) +
  geom_line(aes(x = 1:nrow(df), y = USD_to_OMR, color = "Exchange Rate")) +
  labs(title = "Trend of Oil Variables",
       x = "Time",
       y = "Value") +
  scale_color_manual(values=c("blue","red","orange","black"))
```

# **EDA**

```{r}

data <- df_imputed
str(data)
summary(data)
```

#Histogram of Numerical Variables

```{r}
num_cols <- sapply(data, is.numeric)

par(mfrow=c(2,2))
for(col in names(data)[num_cols]){
  hist(data[[col]], main=col, xlab=col)
}

# Visualize the distribution of oil prices
ggplot(data, aes(x = prices)) + 
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Oil Prices", x = "Oil Prices", y = "Frequency")

# Visualize the distribution of exports
ggplot(data, aes(x = exports)) + 
  geom_histogram(binwidth = 1000, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Oil Exports", x = "Exports (USD)", y = "Frequency")

# Visualize the distribution of production
ggplot(data, aes(x = productions)) + 
  geom_histogram(binwidth = 1000, fill = "lightcoral", color = "black") +
  labs(title = "Distribution of Oil Production", x = "Production", y = "Frequency")

# Visualize the trends of oil prices, exports, and production over time
ggplot(data, aes(x = date)) +
  
  geom_line(aes(y = exports, color = "Oil Exports")) +
  geom_line(aes(y = productions, color = "Oil Production")) +
  labs(title = "Trends of Oil Exports, and Production Over Time", x = "Date", y = "Value") +
  scale_color_manual(values = c("blue", "orange"))


```

```{r}
# Visualize the distribution of oil prices with date on the x-axis
ggplot(data, aes(x = date, y = prices)) +
  geom_line(color = "black") +
  labs(title = "Distribution of Oil Prices Over Time", x = "Date", y = "Oil Prices")

# Visualize the distribution of exports with date on the x-axis
ggplot(data, aes(x = date, y = exports)) +
  geom_line(color = "red") +
  labs(title = "Distribution of Oil Exports Over Time", x = "Date", y = "Exports (USD)")

# Visualize the distribution of production with date on the x-axis
ggplot(data, aes(x = date, y = productions)) +
  geom_line(color = "lightcoral") +
  labs(title = "Distribution of Oil Production Over Time", x = "Date", y = "Production")

# Visualize the trends of oil prices, exports, and production over time
ggplot(data, aes(x = date)) +
  geom_line(aes(y = prices, color = "Oil Prices")) +
  geom_line(aes(y = exports, color = "Oil Exports")) +
  geom_line(aes(y = productions, color = "Oil Production")) +
  labs(title = "Trends of Oil Prices, Exports, and Production Over Time", x = "Date", y = "Value") +
  scale_color_manual(values = c("blue", "red", "orange"))
```

#box plot check outliers

```{r}
par(mfrow = c(1,2))

# Boxplot for 'prices'
boxplot(data$prices,
        main = "Oil Prices",
        ylab = "Prices (USD)",
        col = "lightblue")

# Boxplot for 'exports'
boxplot(data$exports,
        main = "Oil Exports",
        ylab = "Exports (USD)",
        col = "lightgreen")

# Boxplot for 'production'
boxplot(data$productions,
        main = "Oil production",
        ylab = "oil production",
        col = "red")
```

#outliers

```{r}
# Boxplot to visualize outliers
boxplot(data$prices, main = "Oil Prices (USD)", col = "lightblue")
boxplot(data$exports, main = "Oil Exports (USD)", col = "lightgreen")

# Identify outliers using IQR method
Q1 <- quantile(data$prices, 0.25)
Q3 <- quantile(data$prices, 0.75)
IQR_prices <- Q3 - Q1
lower_bound_prices <- Q1 - 1.5 * IQR_prices
upper_bound_prices <- Q3 + 1.5 * IQR_prices

# Remove outliers for 'prices'
df_no_outliers <- data %>%
  filter(prices >= lower_bound_prices & prices <= upper_bound_prices)

# Visualize boxplot after removing outliers
boxplot(df_no_outliers$prices, main = "Oil Prices (USD) After Removing Outliers", col = "red")


```

# correlations

```{r}

# Load the necessary library
library(corrplot)

# Calculate correlation matrix for the numerical variables
correlation_matrix <- cor(df_imputed[, c("prices", "exports", "productions", "USD_to_OMR")], use = "complete.obs")

# Print the correlation matrix
print(correlation_matrix)

# Plot the correlation matrix
corrplot(correlation_matrix, method = "circle", type = "upper", 
         col = colorRampPalette(c("blue", "white", "red"))(200), 
         title = "Correlation Matrix of Oil Prices, Exports, Production, and USD/OMR")

# Calculate pairwise correlations for specific pairs of interest
cor_prices_exports <- cor(df_imputed$prices, df_imputed$exports, use = "complete.obs")
cor_prices_productions <- cor(df_imputed$prices, df_imputed$productions, use = "complete.obs")
cor_prices_usd_omr <- cor(df_imputed$prices, df_imputed$USD_to_OMR, use = "complete.obs")

# Print pairwise correlations
cat("Correlation between Prices and Exports: ", cor_prices_exports, "\n")
cat("Correlation between Prices and Production: ", cor_prices_productions, "\n")
cat("Correlation between Prices and USD to OMR: ", cor_prices_usd_omr, "\n")
```

```{r}
# Load necessary libraries
library(corrplot)
library(ggplot2)
library(RColorBrewer)

# Calculate correlation matrix for the relevant columns
correlation_matrix <- cor(df_imputed[, c("prices", "exports", "productions", "USD_to_OMR")], use = "complete.obs")

# Create a color palette
col <- colorRampPalette(c("red", "yellow", "green"))(200)

# Plot the correlation matrix
corrplot(correlation_matrix, 
         method = "color",       # Use color for visualization
         type = "upper",         # Show only the upper triangle of the matrix
         col = col,              # Set color palette
         addCoef.col = "black",  # Add the correlation coefficients in black
         tl.col = "black",       # Set color of the text labels
         tl.srt = 45,            # Rotate the labels for better readability
         title = "Correlation Matrix ",
         cex.main = 1.5,         # Adjust title size
         mar = c(0, 0, 3, 0))    # Adjust margins for title space
```

```{r}
ggplot(data, aes(x = date, y = prices)) +
  geom_line(color = "blue") +
  labs(title = "Oil Prices Over Time",
       x = "Date",
       y = "Oil Prices")
```

# One Way ANOVA

Performing One-Way ANOVA Across Different Years

```{r}
# One-Way ANOVA to compare oil prices across months
anova_result <- aov(prices ~ month, data = data)

# Summary of ANOVA result
summary(anova_result)
```

Given the dataset you've shared, we will apply a One-Way ANOVA to test if there are significant differences in oil prices across different years. The year will be the factor variable, and oil prices will be the continuous variable.

Assumptions of One-Way ANOVA: Independence: The observations (years) must be independent of each other. Normality: The oil prices in each group (year) should be approximately normally distributed. Homogeneity of Variances: The variance within each group (year) should be equal. This can be tested using Levene's Test.

The One-Way ANOVA test was conducted to examine whether there are significant differences in oil prices across different levels of oil exports, production, and exchange rate.

```{r}
# Plot oil prices by year using a boxplot
ggplot(df_imputed, aes(x = factor(years), y = prices)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.colour = "red", outlier.shape = 16) +
  labs(title = "Oil Prices by Year", 
       x = "Year", 
       y = "Oil Prices (USD per Barrel)") +
  theme_minimal() +  # Apply a clean theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
```

The results show that the p-value is greater than 0.05 for most variables, indicating that there is **no statistically significant difference** in oil prices across the different groups. This means that variations in oil exports, production, and exchange rate do not significantly affect oil prices when analyzed using ANOVA.

Therefore, we fail to reject the null hypothesis and conclude that there is no significant relationship between oil prices and the selected variables under this method.

# Multiple Linear Regression Modle

```{r}

# Multiple Linear Regression Model
mlr_model <- lm(prices ~ productions + exports, data = data)

# Diagnostic plots for regression model
par(mfrow = c(2, 2))

plot(mlr_model)

# Return plotting area to normal
par(mfrow = c(1, 1))

```


# Log transformation for heteroscedasticity
```{r}
# Log transformation for heteroscedasticity
data$log_prices <- log(data$prices)
data$log_productions <- log(data$productions)
data$log_exports <- log(data$exports)

# New regression model after log transformation
log_model <- lm(log_prices ~ log_productions + log_exports, data = data)

# Summary of the new model
summary(log_model)

# Diagnostic plots after log transformation
par(mfrow = c(2, 2))
plot(log_model)
par(mfrow = c(1, 1))
```


# Scatter Plots with Regression Line

```{r}

ggplot(data, aes(exports, prices)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title="Oil Prices vs Oil Export Values",
       x="Oil Export Values",
       y="Oil Prices")

ggplot(data, aes(productions, prices)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title="Oil Prices vs Domestic Production",
       x="Domestic Oil Production",
       y="Oil Prices")

ggplot(data, aes(USD_to_OMR, prices)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title="Oil Prices vs USD/OMR Exchange Rate",
       x="USD/OMR Exchange Rate",
       y="Oil Prices")
```

Model significance

The multiple regression model examines the relationship between oil prices and three explanatory variables: oil exports, domestic production, and the USD/OMR exchange rate.

Domestic oil production

Domestic oil production shows a positive relationship with oil prices, suggesting that higher production levels tend to be associated with higher oil prices.

Oil export values

Oil export values show a weak relationship with oil prices, indicating that export values may not strongly influence price movements in this dataset.

Exchange rate

The USD/OMR exchange rate shows very little variation, therefore its effect on oil prices is minimal.

Overall model

The regression model helps explain the variation in oil prices using the selected economic variables, although the strength of the relationships varies among predictors.

```{r}
library(car)
# run regression model
model <- lm(prices ~ exports + productions + USD_to_OMR, data=data)
# VIF test
vif(model)
```

The Variance Inflation Factor (VIF) was used to check for multicollinearity among the independent variables. The VIF values for exports (1.21), production (1.21), and exchange rate (1.004) are all close to 1 and well below the threshold value of 5.

This indicates that there is **no multicollinearity problem** among the variables. Therefore, the independent variables are not highly correlated with each other, and the regression model is reliable.

# Regression Results

```{r}

library(stargazer)

stargazer(model,
          type = "text",
          title = "Regression Results: Determinants of Oil Prices",
          dep.var.labels = "Oil Prices",
          covariate.labels = c("Oil Export Values",
                               "Domestic Oil Production",
                               "USD/OMR Exchange Rate"),
          digits = 3)
```

The regression analysis was conducted to examine the effect of oil export values, domestic oil production, and the USD/OMR exchange rate on oil prices in Oman.

The overall model is statistically significant (**F = 17.637, p \< 0.01**), indicating that the independent variables jointly have a significant effect on oil prices. The model explains about **39.8% of the variation** in oil prices (**R² = 0.398**), which shows a moderate level of explanatory power.

Among the independent variables, **domestic oil production** has a **positive and highly significant effect** on oil prices (**p \< 0.01**). This means that an increase in production is associated with an increase in oil prices.

**Oil export values** have a **negative and statistically significant effect** (**p \< 0.05**), indicating that higher export levels are associated with lower oil prices.

However, the **USD/OMR exchange rate** is **not statistically significant**, suggesting that it does not have a meaningful impact on oil prices in this model.

In conclusion, oil prices in Oman are mainly influenced by production and export levels, while the exchange rate has no significant effect.

# Durbin Watson test

```{r}
library(lmtest)
# Durbin Watson test
dwtest(model)
```

The Durbin-Watson test was used to check for autocorrelation in the residuals of the regression model. The test result shows a Durbin-Watson value of **0.511**, which is far below 2, and a **p-value less than 0.001**.

This indicates that there is **significant positive autocorrelation** in the residuals. Therefore, the assumption of independent errors is violated.

In conclusion, the regression model has autocorrelation, which may affect the reliability of the results.

# hypothesis testing########################
```{r}
library(ggplot2)
library(dplyr)

# Filter 2016 data
data_2016 <- data %>%
  filter(years == 2016)

# Regression model for 2016
model_2016 <- lm(prices ~ productions + exports, data = data_2016)

# Summary
summary(model_2016)

# ANOVA table for F-test
anova(model_2016)

# Plot 1: Effect of Production on Price
ggplot(data_2016, aes(x = productions, y = prices)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Effect of Production on Price in 2016",
       x = "Production",
       y = "Oil Prices") +
  theme_minimal()

# Plot 2: Effect of Exports on Price
ggplot(data_2016, aes(x = exports, y = prices)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Effect of Exports on Price in 2016",
       x = "Exports",
       y = "Oil Prices") +
  theme_minimal()
```

```{r}
# Filter data for 2017–2018
data_2017_2018 <- data %>%
  filter(years %in% c(2017, 2018))

# Multiple Linear Regression for 2017–2018
model_2017_2018 <- lm(prices ~ productions + exports, data = data_2017_2018)

# Model summary (t-test)
summary(model_2017_2018)

# ANOVA table (F-test)
anova(model_2017_2018)

# Plot 1: Effect of Production on Price
ggplot(data_2017_2018, aes(x = productions, y = prices)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Effect of Production on Price for 2017–2018",
       x = "Production",
       y = "Oil Prices") +
  theme_minimal()

# Plot 2: Effect of Exports on Price
ggplot(data_2017_2018, aes(x = exports, y = prices)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Effect of Exports on Price for 2017–2018",
       x = "Exports",
       y = "Oil Prices") +
  theme_minimal()
```


```{r}
library(dplyr)
library(ggplot2)

# Filter data for 2019–2021
data_2019_2021 <- data %>%
  filter(years %in% c(2019, 2020, 2021))

# Multiple Linear Regression
model_2019_2021 <- lm(prices ~ productions + exports, data = data_2019_2021)

# Summary (t-test)
summary(model_2019_2021)

# ANOVA (F-test)
anova(model_2019_2021)

# Plot 1: Production vs Price
ggplot(data_2019_2021, aes(x = productions, y = prices)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Effect of Production on Price for 2019–2021",
       x = "Production",
       y = "Oil Prices") +
  theme_minimal()

# Plot 2: Exports vs Price
ggplot(data_2019_2021, aes(x = exports, y = prices)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Effect of Exports on Price for 2019–2021",
       x = "Exports",
       y = "Oil Prices") +
  theme_minimal()
```


```{r}
library(dplyr)
library(ggplot2)

# Filter data for 2022
data_2022 <- data %>%
  filter(years == 2022)

# Multiple Linear Regression for 2022
model_2022 <- lm(prices ~ productions + exports, data = data_2022)

# Summary for t-test
summary(model_2022)

# ANOVA table for F-test
anova(model_2022)

# Plot 1: Effect of Production on Price for 2022
ggplot(data_2022, aes(x = productions, y = prices)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Effect of Production on Price for 2022",
       x = "Production",
       y = "Oil Prices") +
  theme_minimal()

# Plot 2: Effect of Exports on Price for 2022
ggplot(data_2022, aes(x = exports, y = prices)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Effect of Exports on Price for 2022",
       x = "Exports",
       y = "Oil Prices") +
  theme_minimal()
```

```{r}
# 3.4 Confidence Intervals in Multiple Regression

# Multiple Linear Regression model
model_full <- lm(prices ~ productions + exports, data = data)

# Model summary
summary(model_full)

# 95% confidence intervals for regression coefficients
confint(model_full, level = 0.95)

# Put coefficients and confidence intervals in one table
coef_table <- cbind(
  Estimate = coef(model_full),
  confint(model_full, level = 0.95)
)

coef_table
```

```{r}
# Confidence intervals for 2016
confint(model_2016, level = 0.95)

# Confidence intervals for 2017–2018
confint(model_2017_2018, level = 0.95)

# Confidence intervals for 2019–2021
confint(model_2019_2021, level = 0.95)

# Confidence intervals for 2022
confint(model_2022, level = 0.95)
```



```{r}
# Fit regression for each period
# 2016
data_2016 <- data %>% filter(years == 2016)
model_2016 <- lm(prices ~ exports + productions + USD_to_OMR, data = data_2016)
summary(model_2016)

# 2017-2018
data_2017_2018 <- data %>% filter(years >= 2017 & years <= 2018)
model_2017_2018 <- lm(prices ~ exports + productions + USD_to_OMR, data = data_2017_2018)
summary(model_2017_2018)

# 2019-2021
data_2019_2021 <- data %>% filter(years >= 2019 & years <= 2021)
model_2019_2021 <- lm(prices ~ exports + productions + USD_to_OMR, data = data_2019_2021)
summary(model_2019_2021)

# 2022
data_2022 <- data %>% filter(years == 2022)
model_2022 <- lm(prices ~ exports + productions + USD_to_OMR, data = data_2022)
summary(model_2022)


```

## 

# Confidence Intervals

```{r}
# Store confidence intervals
ci_2016 <- confint(model_2016)
ci_2017_2018 <- confint(model_2017_2018)
ci_2019_2021 <- confint(model_2019_2021)
ci_2022 <- confint(model_2022)

# Print results
ci_2016
ci_2017_2018
ci_2019_2021
ci_2022
```

prices \~ exports + productions + USD_to_OMR, data


### Time Series
```{r}
# Convert to time series (monthly data)
ts_prices <- ts(data$prices, start = c(2016, 1), frequency = 12)

# Plot
plot(ts_prices,
     type = "l",
     col = "black",
     lwd = 2,
     main = "Oil Prices Over Time",
     xlab = "Time",
     ylab = "Prices")

# Add grid (optional like your figure style)
grid()
```
```{r}
# Plot SAC (ACF)
acf(ts_prices, 
    main = "SAC (ACF) of Oil Prices")
```

```{r}
# First differencing
diff_prices <- diff(ts_prices,lag=1)

# Plot differenced series
plot(diff_prices,
     type = "l",
     col = "blue",
     main = "transformed Oil Prices",
     xlab = "Time",
     ylab = "Differenced Prices")

# SAC after differencing
acf(diff_prices, 
    main = "SAC after Differencing")
```
# Fit ARIMA(0,1,1)

```{r}
library(forecast)

# Convert to time series
ts_prices <- ts(data$prices, start = c(2016,1), frequency = 12)

# Fit ARIMA(0,1,1)
model_arima <- Arima(ts_prices, order = c(0,1,1))

# Summary
summary(model_arima)
auto_model <- auto.arima(ts_prices)

summary(auto_model)
```
#Forecast of Oil Prices
```{r}
# Fit ARIMA(0,1,1)
fitprice <- Arima(ts_prices, order = c(0, 1, 1))

# Forecast next 12 months
price_forecast <- forecast(fitprice, h = 12)

# Print forecast values
print(price_forecast)

# Plot forecast
plot(price_forecast,
     main = "Forecast of Oil Prices",
     xlab = "Time",
     ylab = "Oil Prices")

autoplot(price_forecast) +
  labs(title = "Forecast of Oil Prices",
       x = "Time",
       y = "Oil Prices") +
  theme_minimal()
```






















```{r}
par(mfcol=c(1,2))
plot(data$prices, type="l")
plot(data$productions,type="l")
acf(data$prices)
acf(data$productions)


dprice<-diff(data$prices,lag=1)

par(mfcol=c(1,2))
acf(dprice)
pacf(dprice)
```

### Time Series Plot

```{r}
fitprice<-arima(ts(data$prices),order=c(1,1,0))
fitprice
resprice<-fitprice$residuals
fitsta<-arima(ts(dprice),order=c(1,0,0))
fitsta
par(mfrow=c(1,2))
acf(resprice)
pacf(resprice)

```

not sig \### adequacy model

```{r}
Box.test(fitprice$residuals, lag=1,type = c("Box-Pierce", "Ljung-Box"), fitdf = 0)

```

```{r}
library(forecast)
prediction<-forecast(fitprice,h=15)
plot(prediction)
prediction
```

```{r}
############### Estimation of the parameters SVR
Price<-data$prices
y<-ts(Price)
T<-length(y)
M<-floor(T/4)
y1<-y[1:(T-M)]
fit.price<-arima( y1 , order=c(1,1,0))
summary(fit.price)
fitted<-forecast(fit.price)$fitted
#prediction_price<-forecast(fit.price, h=6)
#prediction_price
#win.graph(width=5, height=5,pointsize=8)
#par(mfrow=c(1,1))
#plot(prediction_price)
#attributes(forecast(fit.price)
```

```{r}

######################### SVR method
const<-rep(1, times=84)
ytrain=y[2:(T-M)]
ytest=y[(T-M+1):T]
xtrain=cbind(const[2:(T-M)],y[1:(T-M-1)])
xtest=cbind(const[(T-M+1):T],y[(T-M):(T-1)])

```

```{r}
library(e1071)
tobj<-tune.svm(x=xtrain,y=ytrain,cost=c(2,5,25,50),epsilon=c(0.005,0.01,.025,.1,0.25))
tobj$best.parameters
Cvalue=tobj$best.parameters$cost
epsvalue=tobj$best.parameters$epsilon


```

```{r}
# Train final SVR model
model_svr <- svm(xtrain, ytrain, cost = 25, epsilon = 0.25)

# Prediction
pred_svr <- predict(model_svr, xtest)

# Accuracy
library(Metrics)
rmse(ytest, pred_svr)
mae(ytest, pred_svr)
```
```{r}
# Prepare SVR data without constant column
xtrain <- matrix(y[1:(T-M-1)], ncol = 1)
xtest  <- matrix(y[(T-M):(T-1)], ncol = 1)

ytrain <- y[2:(T-M)]
ytest  <- y[(T-M+1):T]

# Tune SVR model
library(e1071)

tobj <- tune.svm(
  x = xtrain,
  y = ytrain,
  cost = c(2, 5, 25, 50),
  epsilon = c(0.005, 0.01, 0.025, 0.1, 0.25)
)

tobj$best.parameters

# Train final SVR model
model_svr <- svm(xtrain, ytrain, cost = 25, epsilon = 0.25)

# Prediction
pred_svr <- predict(model_svr, xtest)

# Accuracy
library(Metrics)

rmse(ytest, pred_svr)
mae(ytest, pred_svr)
```





```{r}
library(kernlab)
#Fit ARIMA-SVR with Gaussian,, linear and polynomial kernel respectively 
#validation
regm<-ksvm(xtrain,ytrain,epsilon=epsvalue,C=Cvalue,cross=5)
regmlin<-ksvm(xtrain,ytrain,epsilon=epsvalue,kernel="vanilladot",C=Cvalue,cross=5)
regmpoly<-ksvm(xtrain,ytrain,epsilon=epsvalue,kernel="polydot",C=Cvalue,cross=5)


```

```{r}
predarima<-forecast(fit.price,h=M)
hatarima<-ts(predarima$mean)
attributes(hatarima)
```

```{r}
library(Metrics)
bestgarchmse1=c(rmse(hatarima,ytest),
                rmse(predict(regm,xtest),ytest),
                rmse(predict(regmlin,xtest),ytest),
                rmse(predict(regmpoly,xtest),ytest))
bestgarchmse1

```
```{r}
svrarima<-predict(regmlin,xtest)
```

```{r}
#par(mfrow = c(1, 1))
#win.graph(width=6, height=5,pointsize=8)
plot(y[(T-M+1):T],type="l",ylim=range(0,max(y)+10),col="blue",main="original",xlab="Date(Daily)",ylab="",lty=1)
lines(hatarima,type="l",col="red",main="arimaforecast",xlab="Date(Daily)",ylab="",lty=1)
lines(svrarima,type="l",col="green",main="svm forecast",xlab="Date(Daily)",ylab="",lty=1)
legend(x=2,y=max(y)+5,c("original","arima","SVM_arima"),
       cex=.9,col=c("blue","red","green"),pch="-",lty=5,lwd=2,box.col = "white")


```
